---
title: 'HW #2'
author: "Orla O'Brien"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Load in libraries
library(Sleuth3)
library(Sleuth2)
library(tidyverse)
library(cowplot)
library(kableExtra)
library(mosaic)
```




# Question 2.22: Male vs Female Intelligence

First load in and name the data.
```{r}
dat22 <- Sleuth3::ex0222
head(dat22)
```

Lets take a look at a box plot:
```{r}
ggplot(data=dat22, aes(Gender, AFQT)) +
  geom_boxplot()

```

The way the data is formatted right now, it's hard to plot all the different scores together. So I'm going to reformat using pivot_longer to create one column with all the scores, and then replot the boxplots. 

```{r}
dat22 %>%
  pivot_longer(cols=2:6, names_to="Test", values_to = "Score") %>%

ggplot(aes(Gender, Score, fill=Gender)) +
  geom_boxplot()+
  facet_wrap(~Test) +
  theme_bw() +
  scale_fill_manual(breaks=c("female", "male"), 
                    values=c("lavender", "#99d8c9")) +
  scale_x_discrete(labels=c("Female", "Male")) +
  labs(title="Composite and Individual AFQT Test Scores by Gender")

```

Now I want to perform my t-tests and show them in a table.

```{r}

tests <- list()


tests[[1]] <- t.test(AFQT ~ Gender, var.equal = TRUE, data = dat22)
tests[[2]] <-t.test(Arith ~ Gender, var.equal = TRUE, data = dat22)
tests[[3]] <-t.test(Math ~ Gender, var.equal = TRUE, data = dat22)
tests[[4]] <-t.test(Parag ~ Gender, var.equal = TRUE, data = dat22)
tests[[5]] <-t.test(Word ~ Gender, var.equal = TRUE, data = dat22)

results <- sapply(tests, function(x) {
  c(x$estimate[1],
    x$estimate[2],
    ci.lower = x$conf.int[1],
    ci.upper = x$conf.int[2],
    p.value = x$p.value)
})
colnames(results) <- c("AFQT", "Arithmetic", "Math", "Word Knowledge", "Paragraph Comprehension")
rownames(results) <- c("Mean Score (Female)", "Mean Score (Male)", "Confidence Interval (Upper)",
                       "Confidence Interval (Lower)", "p-value")

results
 kbl(results, booktabs = T, longtable = T, caption = "Statistical Summary", linesep = "") %>%
 kable_styling(latex_options = c("hold_position", "repeat_header"), full_width = F, font_size = 9)
```



# Question 3.32: College Tuition 

Load in and name the data

```{r}
dat32 <- Sleuth3::ex0332 
head(dat32)
```


I want to make sure that I can plot things easily, so I'll pivot the data to one column.


```{r}
dat32_pivot <- dat32 %>% pivot_longer(cols=3:4, names_to="InOut", values_to = "Cost")
```

Now, take a look at the spread of the data
```{r}
dat32_pivot %>%
  filter(Type=="Public") %>%
  ggplot(aes(InOut, Cost, fill=InOut)) +
  geom_boxplot() +
  theme_bw() +
  scale_fill_manual(breaks=c("InState", "OutOfState"), 
                    values=c("#d95f02", "#1b9e77")) +
  scale_x_discrete(labels=c("In State Tuition", "Out of State Tuition")) +
  labs(x= "", y="Cost (USD)", title="Comparison of College Tuition")
```

The boxplot shows that the interquartile ranges are quite different. However, I had a hard time visualizing this so I plotted it as a histogram as well to look at the spread.

```{r}
dat32_pivot %>%
  ggplot(aes(Cost)) +
  geom_histogram(binwidth=2000, color="black", fill="blue") +
  facet_grid(rows=vars(Type), cols=vars(InOut)) +
  theme_bw()+
  labs(y="Count", x="Cost (USD)")
```

Data are clearly not normal, so we need to do a transformation.

I'll use BoxCox to see what kind of transformation I need to do.


```{r}
MASS::boxcox(Cost ~ Type, data=dat32_pivot,lam=seq(-1,1,.1))

```


It looks like a log transformation would be appropriate, because lambda is zero.

Now I'll make box plots for the three comparisons the question asks for. I'm using cowplot to plot them together.

```{r}

In.V.Out <- dat32_pivot %>%
  filter(Type=="Public") %>%
  ggplot(aes(InOut, log(Cost), fill=InOut)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position="none", plot.title = element_text(size=10)) +
  scale_fill_manual(breaks=c("InState", "OutOfState"), 
                    values=c("#d95f02", "#1b9e77")) +
  scale_x_discrete(labels=c("In State Tuition", "Out of State Tuition")) +
labs(x= "", y="Log of Cost (USD)", title="In vs Out of State Public College Tuition")

Private.V.Public <- dat32_pivot %>%
  filter(InOut=="InState") %>%
  ggplot(aes(Type, log(Cost), fill=Type)) +
  geom_boxplot() + 
  theme_bw() +
  theme(legend.position="none", plot.title = element_text(size=10)) +
  scale_fill_manual(breaks=c("Private", "Public"), 
                    values=c("#d95f02", "#1b9e77")) +
  scale_x_discrete(labels=c("In State Private", "In State Public")) +
  labs(x= "", y="Log of Cost (USD)", title="Private vs Public College In-State Tuition")

Private.V.PublicOut <- dat32_pivot %>%
  filter(InOut=="OutOfState") %>%
  ggplot(aes(Type, log(Cost), fill=Type)) +
  geom_boxplot() + 
  theme_bw() +
  theme(legend.position="none", plot.title = element_text(size=10)) +
  scale_fill_manual(breaks=c("Private", "Public"), 
                    values=c("#d95f02", "#1b9e77")) +
  scale_x_discrete(labels=c("Out of State Private", "Out of State Public")) +
  labs(x= "", y="Log of Cost (USD)", title="Private vs Public College Out of State Tuition")

plot_grid(In.V.Out, Private.V.Public, Private.V.PublicOut, labels = c('A', 'B', 'C'))
```

Now we want to look at the statistical summary for these three comparisons.

```{r}

tests32 <- list()
t1 <- dat32_pivot %>% filter(Type=="Public") 
t3 <- dat32_pivot %>% filter(InOut=="OutOfState")


tests32[[1]] <- t.test(log(Cost) ~ InOut, data=t1)
tests32[[2]] <- t.test(log(InState) ~ Type, data=dat32)
tests32[[3]] <- t.test(log(Cost) ~ Type, data=t3)

results32 <- sapply(tests32, function(x) {
  c(x$estimate[1],
    x$estimate[2],
    ci.lower = x$conf.int[1],
    ci.upper = x$conf.int[2],
    p.value = x$p.value)
})

colnames(results32) <- c("In Vs Out of State", "Priv vs Pub (In State)","Priv vs Pub (Out Of State)" )
rownames(results32) <- c("Mean Score (Group 1)", "Mean Score (Group 2)", "Confidence Interval (Upper)",
                       "Confidence Interval (Lower)", "p-value")

 kbl(results32, booktabs = T, longtable = T, caption = "Statistical Summary", linesep = "") %>%
 kable_styling(latex_options = c("hold_position", "repeat_header"), full_width = F, font_size = 9)
```

The results show that there is strong evidence that out of state public school tuition is higher than in state public school tuition, and that private school tuition is higher than public school tuition when looking at either in or out of state costs.

# Question 2.21 from Sleuth 2: Weights of Birds that Survived and Perished 

Read in and check data.
```{r}
dat0221 <- Sleuth2::ex0221
head(dat0221)
```

```{r}

ggplot(dat0221, mapping=aes(Status, Weight, fill=Status)) +
  geom_boxplot()+
  theme_bw() +
  theme(legend.position="none") +
  scale_fill_manual(breaks=c("perished", "survived"), 
                    values=c("#d7301f", "#fdbb84")) +
  scale_x_discrete(labels=c("Perished", "Survived")) +
  labs(x= "", y="Sparrow Weight", title="Weights of Sparrows that Perished or Survived During a Winter Storm")


result21 <- favstats(Weight ~ Status, data=dat0221)


tests21 <- list()

tests21[[1]] <- t.test(Weight ~ Status, data=dat0221)

Tresults21 <- sapply(tests21, function(x) {
  c(x$estimate[1],
    x$estimate[2],
    ci.lower = x$conf.int[1],
    ci.upper = x$conf.int[2],
    p.value = x$p.value)
})

kbl(result21, booktabs = T, longtable = T, caption = "Statistical Summary", linesep = "") %>%
 kable_styling(latex_options = c("hold_position", "repeat_header"), full_width = F, font_size = 9)

kbl(Tresults21, booktabs = T, longtable = T, caption = "Statistical Summary", linesep = "") %>%
 kable_styling(latex_options = c("hold_position", "repeat_header"), full_width = F, font_size = 9)
```



# Question 3.31 from Sleuth 2: Iron Supplementation

Load in and check the data
```{r}
dat0331 <- Sleuth2::ex0331
head(dat0331)
```

Check the spread to see if we need to transform
```{r}
a <- ggplot(dat0331, mapping=aes(Supplement,Iron, fill=Supplement)) +
  geom_boxplot()+
  scale_fill_manual(values=c("Fe3" = "#238b45", "Fe4" = "#4292c6")) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(y= "Percent Iron Retained", x="")

b <- ggplot(dat0331, mapping=aes(Iron, fill=Supplement, linetype = Supplement, color=Supplement)) +
  geom_histogram(position="identity", alpha=.5, binwidth = .5) +
  scale_fill_manual(values=c("Fe3" = "#238b45", "Fe4" = "#4292c6")) +
  scale_color_manual(values=c("Fe3" = "#238b45", "Fe4" = "#4292c6")) + 
  theme_bw() +
  theme(legend.position = "none") +
  labs(x= "Percent Iron Retained", y="Count")

plot_grid(a, b, ncol=1, labels = c('A', 'B'))
```



There are some outliers here that maybe we should get rid of.
```{r}
dat0331 <- subset(dat0331, (Supplement=="Fe3" & Iron < 7.5) | (Supplement=="Fe4" & Iron < 10))

a <- ggplot(dat0331, mapping=aes(Supplement,Iron, fill=Supplement)) +
  geom_boxplot()+
  scale_fill_manual(values=c("Fe3" = "#238b45", "Fe4" = "#4292c6")) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(y= "Percent Iron Retained", x="")

b <- ggplot(dat0331, mapping=aes(Iron, fill=Supplement, linetype = Supplement, color=Supplement)) +
  geom_histogram(position="identity", alpha=.5, binwidth = .5) +
  scale_fill_manual(values=c("Fe3" = "#238b45", "Fe4" = "#4292c6")) +
  scale_color_manual(values=c("Fe3" = "#238b45", "Fe4" = "#4292c6")) + 
  theme_bw() +
  theme(legend.position = "none") +
  labs(x= "Percent Iron Retained", y="Count")

plot_grid(a, b, ncol=1, labels = c('A', 'B'))
```

This looks better, but I still think we need a transformation. I'll use BoxCox to see what type of transformation is warranted.

```{r}
MASS::boxcox(Iron ~ Supplement, data=dat0331,lam=seq(-1,2,.1))
```

Looks like a square root transformation is suggested (lambda is roughly 0.5).
Now we'll view the data again, do our T test, and report the statistical summary.
..
```{r}
ggplot(dat0331, mapping=aes(Supplement,sqrt(Iron), fill=Supplement)) +
  geom_boxplot()+
  scale_fill_manual(values=c("Fe3" = "#238b45", "Fe4" = "#4292c6")) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(y= "Percent Iron Retained (Square Root Transformed)", x="")
```

```{r}
t.test(sqrt(Iron) ~ Supplement,  data=dat0331)

results31 <- favstats(sqrt(Iron) ~ Supplement,  data=dat0331)




tests31 <- list()

tests31[[1]] <- t.test(sqrt(Iron) ~ Supplement,  data=dat0331)

Tresults31 <- sapply(tests31, function(x) {
  c(x$estimate[1],
    x$estimate[2],
    ci.lower = x$conf.int[1],
    ci.upper = x$conf.int[2],
    p.value = x$p.value)
})

kbl(results31, booktabs = T, longtable = T, caption = "Statistical Summary", linesep = "") %>%
 kable_styling(latex_options = c("hold_position", "repeat_header"), full_width = F, font_size = 9)

kbl(Tresults31, booktabs = T, longtable = T, caption = "Statistical Summary", linesep = "") %>%
 kable_styling(latex_options = c("hold_position", "repeat_header"), full_width = F, font_size = 9)
```













